% LaTeX source for textbook ``How to think like a computer scientist''
% Copyright (C) 1999  Allen B. Downey

% This LaTeX source is free software; you can redistribute it and/or
% modify it under the terms of the GNU General Public License as
% published by the Free Software Foundation (version 2).

% This LaTeX source is distributed in the hope that it will be useful,
% but WITHOUT ANY WARRANTY; without even the implied warranty of
% MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
% General Public License for more details.

% Compiling this LaTeX source has the effect of generating
% a device-independent representation of a textbook, which
% can be converted to other formats and printed.  All intermediate
% representations (including DVI and Postscript), and all printed
% copies of the textbook are also covered by the GNU General
% Public License.

% This distribution includes a file named COPYING that contains the text
% of the GNU General Public License.  If it is missing, you can obtain
% it from www.gnu.org or by writing to the Free Software Foundation,
% Inc., 59 Temple Place - Suite 330, Boston, MA 02111-1307, USA.


\documentclass{book}
\usepackage{epsfig}
\usepackage{makeidx}
\pssilent

\title{How to think like a computer scientist}

\author{Allen B. Downey}
\date{}

\sloppy
\setlength{\topmargin}{0.75in}
\setlength{\headsep}{0.5in}
\setlength{\oddsidemargin}{1.0in}
\setlength{\evensidemargin}{.95in}
\makeindex

\begin{document}

\title {How to think like a computer scientist}
\author {Allen B. Downey}
\date {C++ Version, First Edition}
\maketitle

\vspace{2in}
\begin{center}
{\Large How to think like a computer scientist}

C++ Version, First Edition
\vspace{0.25in}

Copyright (C) 1999  Allen B. Downey
\end{center}
\vspace{0.25in}

This book is an Open Source Textbook (OST).  Permission is
granted to reproduce, store or transmit the text of this
book by any means, electrical, mechanical, or biological,
in accordance with the terms of the GNU General Public License as
published by the Free Software Foundation (version 2).

This book is distributed in the hope that it will be useful,
but WITHOUT ANY WARRANTY; without even the implied warranty of
MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
General Public License for more details.

The original form of this book is LaTeX source code.
Compiling this LaTeX source has the effect of generating
a device-independent representation of a textbook, which
can be converted to other formats and printed.  All intermediate
representations (including DVI and Postscript), and all printed
copies of the textbook are also covered by the GNU General
Public License.

The LaTeX source for this book, and more information about
the Open Source Textbook project, is available from

\begin{verbatim}
     http://www.cs.colby.edu/~downey/ost
\end{verbatim}
%
or by writing to Allen B. Downey, 5850 Mayflower Hill,
Waterville, ME 04901.

The GNU General Public License is available from
www.gnu.org or by writing to the Free Software Foundation,
Inc., 59 Temple Place - Suite 330, Boston, MA 02111-1307, USA.

This book was typeset by the author using LaTeX and dvips,
which are both free, open-source programs.


\frontmatter
\tableofcontents

\mainmatter
\include{cha1}
\include{cha2}
\include{cha3}
\include{cha4}
\include{cha5}
\include{cha6}
\include{cha7}
\include{cha8}
\include{cha9}
\include{cha10}
\include{cha11}
\include{cha12}
\include{cha13}
\include{cha14}
\include{cha15}

\appendix
\include{app1}

\printindex

\end{document}



